Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics

### 2017-09-22 Manuscripts_GMD_ap_ref1.tex
done proof reading

Plume-SPH provides the first particle based simulation of volcanic plumes. SPH (smoothed particle hydrodynamics) has several advantages over currently used mesh based methods in modeling of multiphase free boundary flows like volcanic plumes. This tool will provide more accurate eruption source terms to users of VATDs (Volcanic ash transport and dispersion models) greatly improving volcanic ash forecasts. The accuracy of these terms is crucial for forecasts from VATDs and the 3D SPH model presented here will provide better numerical accuracy. As an initial effort to exploit the feasibility and advantages of SPH in volcanic plume modeling, we adopt a relatively simple physics model (3D dusty-gas dynamic model assuming well mixed eruption material, dynamic equilibrium and thermodynamic equilibrium between erupted material and air that entrained into the plume and minimal effect of winds) targeted at capturing the salient features of a volcanic plume. The documented open source code is easily obtained and extended to incorporate other models of physics of interest to the large community of researchers investigating multiphase free boundary flows of volcanic or other origins.

@@

The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak and strong scalability using novel techniques for data management using a SFCs (space-filling curves) and object creation time based indexing and hash table based storage scheme. These techniques are of interest to  researchers engaged in developing particle in cell type methods. The code is first verified by 1D shock tube tests, then by comparing velocity and concentration distribution along the central axis and on the transverse
cross with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). 
%The model is validated by comparing against simulation results from other 3D plume models.
Profiles of several integrated variables are compared with those calculated by existing 3D plume models for an eruption with the same MER (mass eruption rate) estimated for the Pinatubo eruption of June 15 1991. Our results are consistent with existing 3D plume models. Analysis of the plume evolution process illustrates that this model is able to reproduce the physics of plume development.

@@

While SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM (Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, dynamic and thermodynamical equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} captures turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules, including gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiple phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and computationally efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step are used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adopted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing large process counts. The core solver of our model is parallelized by distributed memory MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) induced indexing and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

\subsection{Boundary conditions}
In the current model the initial domain is a box. The boundaries are categorized as the velocity inlet (a circular area at the center of the bottom of the box), wall boundary (box bottom), pressure outlet (other faces of the box).
%probably should add a section to list the input parameters in the verification section

@@

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function $\delta(\textbf{x})$, for an arbitary function $A(\textbf{x})$, the following equation holds. 
\begin{equation}
A\left(\textbf{x}\right)=\int_{-\infty}^{\infty} A\left(\textbf{x} \prime\right) \delta \left(\textbf{x} \prime - \textbf{x}\right) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w\left(\textbf{x}-\textbf{x}\prime, h\right)$ (or $w\left(\textbf{x}\prime-\textbf{x}, h\right)$) which tends to a Dirac function when the smoothing length $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w\left(\textbf{x} \prime-\textbf{x}, h\right) =  \delta \left(\textbf{x} \prime - \textbf{x}\right)
\label{eq:SPH_kernel_delta}
\end{equation}
So it can be viewed as an approximate form of the Dirac function and should satisfy the normalization condition:
\begin{equation}
\int	 w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime = 1
\label{eq:SPH-kernel-normalization-prop}
\end{equation}
Besides normalization, the weighting function of particle $a$ has to be symmetric with respect to $a$ to ensure that neighbor particles located at the same distance away from $a$ contribute equally to SPH summation equation, see Eq. (\ref{eq:SPH-kernel-symmetric}).
\begin{equation}
w\left(\textbf{x}- \textbf{x} \prime, h\right) = w\left(\textbf{x} \prime - \textbf{x}, h\right)
\label{eq:SPH-kernel-symmetric}
\end{equation}
The weighting function also needs to satisfy conditions such as positivity and compact support. In addition, the kernel function must be monotonically decreasing with the distance between particles.

@@

The SPH viscosity can be related to a continuum viscosity by converting the summation to integrals \citep{monaghan2005smoothed}. It has been shown that  shear viscosity coefficient $\eta = \frac{\rho \alpha h c}{8} $ and bulk viscosity coefficient $ \zeta = \frac{5 \eta}{3}$ are appropriate for two dimensional flows and modified to $\eta = \frac{\rho \alpha h c}{10} $ , $ \zeta = \frac{5 \eta}{3}$ for three dimensional flows.
An extra term was added to $\nu$ considering aspects of the dissipative term in shock solutions based on Riemann solvers and lead to a new formulation of artificial viscosity. We adopt this new formulation in our simulation:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
$\alpha$ and $\beta$ are two parameters that can be adjusted for different cases.
$\alpha = 1$ and $\beta = 2$ are  recommended by Monaghan for best results. In our simulation, these two parameters are calibrated to  $\alpha = 0.3$ and $\beta = 0.6$. $\eta$ is usually taken as 0.1 to prevent singularities.
%when $\textbf{x}_{ab} = 0$.

\subsection{Discretization of governing equations and extensibility}
The basic interpolation given in Eq. (\ref{eq:SPH-approximation-sum}) to Eq. (\ref{eq:SPH-vecctor-function}) provides a general way to obtain SPH expressions of governing equations. The problem is that using these expressions ``{as is}" in general leads to quite poor gradient estimates. Various tricks can be used to conserve linear and angular momentum and thermal energy \citep{monaghan1992smoothed}. Special treatments are also needed for second order derivative terms \citep{monaghan2005smoothed}. We only refer here to one of these possible discretizations of compressible Euler equations with SPH:
\begin{align}
<\rho_a> = \sum_b m_b w_{ab} \left(h\right) \label{eq:ns-sph-d} \\
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle = -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) +\textbf{g} \label{eq:ns-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-e}
\end{align}
where, $a$ is the SPH particle index, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. $\Pi$ is an artificial viscosity term, which is discussed in section \ref{sec:artificial-viscosity}. $w_{a b}\left(h\right)$ is a concise form of $w\left(\textbf{x}_a - \textbf{x}_b, h\right)$ and from here on, we will use this concise form.
%However, these discretized formulations will not guarantee conservation of entropy. We can obtain conservation of entropy by giving up on conservation of mass or thermal energy adopting  alternative discretized formulations \citep{monaghan1992smoothed}.
As a Lagrangian method, particle position is also updated at every time step.
\begin{equation}
\left\langle\dfrac{d \textbf{x}_a}{dt}\right\rangle = \textbf{v}_a \label{eq:SPH-update-pos}
\end{equation}

@@

\subsection{Time step}
The physical quantities (velocity, density and pressure) and particle position change every time step. The Courant condition, which is in spirit similar to the Courant condition for the mesh-based methods, is used to determine the time step $\Delta t$.
\begin{equation}
\Delta t = \textrm{CFL} \min_a \bigg \lbrace \dfrac{\left[\frac{m_a}{\rho_a}\right]^{\frac{1}{d}}}{c_a} \bigg \rbrace
\end{equation}
where $c_a$ is sound speed at particle $a$ and calculated based on heat specific ration of the mixture $\gamma_m$ (See Eq. (\ref{eq:sound-speed-mixture})). 
$d$ is number of dimensions. 
\begin{equation}
c_a = \left( \gamma_m \frac{p}{\rho} \right)^{0.5}
\label{eq:sound-speed-mixture}
\end{equation}

@@

\subsection{Tensile instability and corrected derivatives}
The classical SPH method was known to suffer from tensile instability and boundary deficiency. Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simple example of such a test calculation exhibiting the instability involves a body which is subjected to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution \citep{swegle1995smoothed}. To address these difficulties, \citep{chen1999improvement} proposed a corrected SPH formulation. For 1D case, employing a Taylor expansion for $A\left(x\right)$ about $x_a$, multiplying both sides by kernel function and then doing integration over the domain gives
\begin{equation}
\int_{\Omega} A\left(x\right) w\left(x- x_a, h\right) dx = 
A_a \int_{\Omega} w\left(x - x_a, h\right) dx +\dfrac{\partial A}{\partial x}(x_a) \int_{\Omega} \left(x-x_a\right) w\left(x - x_a, h\right) dx +...
\end{equation}
Ignoring derivative terms higher than first order, and writing the integral in particle approximation form leads to:
\begin{equation}
A_a = \frac{\sum_b m_b \dfrac{A_b}{\rho_b} w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{1}{\rho_b} w\left(x_a-x_b, h\right)}
\label{eq:CSP-function-approximation-1d}
\end{equation}
Notice that the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) is actually summation approximation of Eq. (\ref{eq:SPH-kernel-normalization-prop}). That is to say, Eq. (\ref{eq:CSP-function-approximation-1d}) and Eq. (\ref{eq:SPH-approximation-sum}) are the same for particles far away from boundaries as the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) becomes $1$ in that case.
The first order derivative term can be obtained in a similar way:
\begin{equation}
\nabla A_a = \frac{\sum_b m_b \dfrac{A_b - A_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{x_b - x_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}
\end{equation}

@@

\subsection{Mass fraction update}
Air and erupted material are represented by two different sets of SPH particles (or discretization points) in the model. Based on assumptions we made in section \ref{sec:physics-model}, only density needs to be updated respectively for each phase. The updating of density is exactly the same as Eq. (\ref{eq:ns-sph-d}) in spirit.  Particles of phase 1 are not be counted while evaluating density of phase 2 and vice versa. Updating of density is then based on the following discretized equations.
\begin{equation}
<\rho_{\alpha}^a>=\frac{\sum m_b w_{\alpha b} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} \left(h\right) +\sum \frac{m_j}{\rho_j} w_{\alpha j} \left(h\right)} \label{eq:gov-sph-d1}
\end{equation}
\begin{equation}
<\rho_\alpha^{sg}>=\frac{\sum_j m_j w_{\alpha j} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} \left(h\right) +\sum \frac{m_j}{\rho_j} w_{\alpha j} \left(h\right)} \label{eq:gov-sph-d2}
\end{equation}
where the subscript $a$ and $b$ represents air particles (phase 1) while $i$ and $j$ represents erupted material. $\beta$ and $\alpha$ represents either erupted material or air.
$\rho_a^a$ is density of phase 1 (air). 
 $\rho_i^{sg}$ is density of phase 2 (erupted material).
$\rho=\rho^a + \rho^{sg}$ is density of mixture of air and erupted material. 
By definition, the mass fraction is updated according to Eq. (\ref{eq:gov-sph-xi}).
\begin{equation}
<\xi_{\alpha}> = \dfrac{\rho^{sg}_{\alpha}}{\rho_{\alpha}}
\label{eq:gov-sph-xi}
\end{equation}

In areas far away from the interface, updating of density is exactly the same as that for single phase flow. For example, in the right side and left side (or blue areas) in Fig. \ref{fig:SPH-multiple-density}, where there are only air particles, Eq.  (\ref{eq:gov-sph-d2}) evaluates to zero and total density is:
\begin{equation}
\rho_{\alpha}=\rho_{\alpha}^a=\frac{\sum m_b w_{\alpha b} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b}}
\end{equation}
Which is a special case of Eq. (\ref{eq:CSP-function-approximation-1d}). For these areas occupied by only particles of phase 2 and far away from the interface, similarly, the equation for density update becomes: 
\begin{equation}
\rho_{\alpha}=\rho_{\alpha}^{sg}=\frac{\sum m_j w_{\alpha j} \left(h\right)}{\sum \frac{m_j}{\rho_j} w_{\alpha j}}
\end{equation}
That is to say, the same density updating equation can be applied for both phases and no additional numerical treatment needed to locate where the interface is.

\begin{figure}
\includegraphics[width=12cm]{Interface}
\caption{In the left figure, the blue particles (phaseID=1) represent air particles the red ones (phaseID=2) represent erupted material. The right figure shows corresponding mass fraction. Mass fraction are evaluated based on Eq. (\ref{eq:gov-sph-d1}) and Eq. (\ref{eq:gov-sph-d1}) without any other interface track or capture method.}
\label{fig:SPH-multiple-density}
\end{figure}

@@

More details about the data structure, domain decomposition, load balancing, domain adjusting and performance testing has been published separately \citep{cao2017data}.

>>>>

Plume-SPH provides the first particle based simulation of volcanic plumes. SPH (smoothed particle hydrodynamics) has several advantages over currently used mesh based methods in modeling of multiphase free boundary flows like volcanic plumes. This tool will provide more accurate eruption source terms to users of VATDs (Volcanic ash transport and dispersion models) greatly improving volcanic ash forecasts. The accuracy of these terms is crucial for forecasts from VATDs and the 3D SPH model presented here will provide better numerical accuracy. As an initial effort to exploit the feasibility and advantages of SPH in volcanic plume modeling, we adopt a relatively simple physics model (3D dusty-gas dynamic model assuming well mixed eruption material, dynamic equilibrium and thermodynamic equilibrium between erupted material and air that entrained into the plume, and minimal effect of winds) targeted at capturing the salient features of a volcanic plume. The documented open source code is easily obtained and extended to incorporate other models of physics of interest to the large community of researchers investigating multiphase free boundary flows of volcanic or other origins.

@@

The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak and strong scalability using novel techniques for data management using a SFCs (space-filling curves) and object creation time based indexing and hash table based storage scheme. These techniques are of interest to  researchers engaged in developing particle in cell type methods. The code is first verified by 1D shock tube tests, then by comparing velocity and concentration distribution along the central axis and on the transverse
cross with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). 
%The model is validated by comparing against simulation results from other 3D plume models.
Profiles of several integrated variables are compared with those calculated by existing 3D plume models for an eruption with the same MER (mass eruption rate) estimated for the Pinatubo eruption of June 15 1991. Our results are consistent with existing 3D plume models. Analysis of the plume evolution process demonstrates that this model is able to reproduce the physics of plume development.

@@

While SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM (Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, dynamic and thermodynamical equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme that differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} is adopted to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules, including gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiple phase flow with less additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a numerically robust and computationally efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step are used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adopted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing using large process counts. The core solver of our model is parallelized by distributed memory MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) induced indexing and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

\subsection{Boundary conditions}
In the current model the initial domain is a box. The boundaries are categorized as the velocity inlet (a circular area at the center of the bottom of the box), wall boundary (box bottom) and pressure outlet (other faces of the box).
%probably should add a section to list the input parameters in the verification section

@@

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function $\delta(\textbf{x})$, for an arbitary function $A(\textbf{x})$, the following equation holds. 
\begin{equation}
A\left(\textbf{x}\right)=\int_{-\infty}^{\infty} A\left(\textbf{x} \prime\right) \delta \left(\textbf{x} \prime - \textbf{x}\right) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function $\delta$ in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w\left(\textbf{x}-\textbf{x}\prime, h\right)$ (or $w\left(\textbf{x}\prime-\textbf{x}, h\right)$) which tends to a Dirac function when the smoothing length $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w\left(\textbf{x} \prime-\textbf{x}, h\right) =  \delta \left(\textbf{x} \prime - \textbf{x}\right)
\label{eq:SPH_kernel_delta}
\end{equation}
The weighting function, as an approximate form of the Dirac function, should satisfy the normalization condition:
\begin{equation}
\int	 w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime = 1
\label{eq:SPH-kernel-normalization-prop}
\end{equation}
Besides normalization, the weighting function of particle $a$ has to be symmetric with respect to $a$ to ensure that neighbor particles located at the same distance away from $a$ contribute equally to SPH summation equation, see Eq. (\ref{eq:SPH-kernel-symmetric}).
\begin{equation}
w\left(\textbf{x}- \textbf{x} \prime, h\right) = w\left(\textbf{x} \prime - \textbf{x}, h\right)
\label{eq:SPH-kernel-symmetric}
\end{equation}
The weighting function also needs to satisfy conditions such as positivity and compact support. In addition, the kernel function must be monotonically decreasing with the distance between particles.

@@

The SPH viscosity can be related to a continuum viscosity by converting the summation to integrals \citep{monaghan2005smoothed}. It has been shown that  shear viscosity coefficient $\eta = \frac{\rho \alpha h c}{8} $ and bulk viscosity coefficient $ \zeta = \frac{5 \eta}{3}$ are appropriate for two dimensional flows. $\eta = \frac{\rho \alpha h c}{10} $ and $ \zeta = \frac{5 \eta}{3}$ are appropriate for three dimensional flows.
An extra term was added to $\nu$ considering aspects of the dissipative term in shock solutions based on Riemann solvers and lead to a new formulation of artificial viscosity. We adopt this new formulation in our simulation:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
$\alpha$ and $\beta$ are two parameters that can be adjusted for different cases.
$\alpha = 1$ and $\beta = 2$ are  recommended by Monaghan for best results. In our simulation, these two parameters are calibrated to  $\alpha = 0.3$ and $\beta = 0.6$. $\eta$ is usually taken as 0.1 to prevent singularities.
%when $\textbf{x}_{ab} = 0$.

\subsection{Discretization of governing equations and extensibility}
The basic interpolation given in Eq. (\ref{eq:SPH-approximation-sum}) to Eq. (\ref{eq:SPH-vecctor-function}) provides a general way to obtain SPH expressions of governing equations. The problem is that using these expressions ``{as is}" in general leads to quite poor gradient estimates. Various tricks can be used to conserve linear and angular momentum and thermal energy \citep{monaghan1992smoothed}. Special treatments are also needed for second order derivative terms \citep{monaghan2005smoothed}. We only refer here to one of these possible discretizations of compressible Euler equations with SPH:
\begin{align}
<\rho_a> = \sum_b m_b w_{ab} \left(h\right) \label{eq:ns-sph-d} \\
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle = -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) +\textbf{g} \label{eq:ns-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-e}
\end{align}
where, $a$ is the SPH particle index. $\Pi$ is an artificial viscosity term, which is discussed in section \ref{sec:artificial-viscosity}. $w_{a b}\left(h\right)$ is a concise form of $w\left(\textbf{x}_a - \textbf{x}_b, h\right)$ and from here on, we will use this concise form.
%However, these discretized formulations will not guarantee conservation of entropy. We can obtain conservation of entropy by giving up on conservation of mass or thermal energy adopting  alternative discretized formulations \citep{monaghan1992smoothed}.
As a Lagrangian method, particle position is also updated at every time step.
\begin{equation}
\left\langle\dfrac{d \textbf{x}_a}{dt}\right\rangle = \textbf{v}_a \label{eq:SPH-update-pos}
\end{equation}

@@

\subsection{Time step}
The physical quantities (velocity, density and pressure) and particle position change every time step. The Courant condition, which is in spirit similar to the Courant condition for the mesh-based methods, is used to determine the time step $\Delta t$.
\begin{equation}
\Delta t = \textrm{CFL} \min_a \bigg \lbrace \dfrac{\left[\frac{m_a}{\rho_a}\right]^{\frac{1}{d}}}{c_a} \bigg \rbrace
\end{equation}
where $c_a$ is sound speed at particle $a$ and calculated based on heat specific ration of the mixture $\gamma_m$ (See Eq. (\ref{eq:sound-speed-mixture})). 
\begin{equation}
c_a = \left( \gamma_m \frac{p}{\rho} \right)^{0.5}
\label{eq:sound-speed-mixture}
\end{equation}

@@

\subsection{Tensile instability and corrected derivatives}
The classical SPH method was known to suffer from tensile instability and boundary deficiency. Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simple example of such a test calculation exhibiting the instability involves a body which is subject to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution \citep{swegle1995smoothed}. To address these difficulties, \citet{chen1999improvement} proposed a corrected SPH formulation. For 1D case, employing a Taylor expansion for $A\left(x\right)$ about $x_a$, multiplying both sides by kernel function and then doing integration over the domain gives
\begin{equation}
\int_{\Omega} A\left(x\right) w\left(x- x_a, h\right) dx = 
A_a \int_{\Omega} w\left(x - x_a, h\right) dx +\dfrac{\partial A}{\partial x}(x_a) \int_{\Omega} \left(x-x_a\right) w\left(x - x_a, h\right) dx +...
\end{equation}
Ignoring derivative terms higher than first order, and writing the integral in particle approximation form leads to:
\begin{equation}
A_a = \frac{\sum_b m_b \dfrac{A_b}{\rho_b} w_{ab}\left(h\right)}{\sum_b m_b \dfrac{1}{\rho_b} w_{ab}\left(h\right)}
\label{eq:CSP-function-approximation-1d}
\end{equation}
Notice that the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) is actually summation approximation of the left side in Eq. (\ref{eq:SPH-kernel-normalization-prop}). That is to say, Eq. (\ref{eq:CSP-function-approximation-1d}) and Eq. (\ref{eq:SPH-approximation-sum}) are the same for particles far away from boundaries as the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) becomes $1$ in that case.
The first order derivative term can be obtained in a similar way:
\begin{equation}
\nabla A_a = \frac{\sum_b m_b \dfrac{A_b - A_a}{\rho_b} \nabla_a w_{ab}\left(h\right)}{\sum_b m_b \dfrac{x_b - x_a}{\rho_b} \nabla_a w_{ab}\left( h\right)}
\end{equation}

@@

\subsection{Mass fraction update}
Air and erupted material are represented by two different sets of SPH particles (or discretization points) in the model. Based on assumptions we made in section \ref{sec:physics-model}, only density needs to be updated respectively for each phase. The updating of density is exactly the same as Eq. (\ref{eq:ns-sph-d}) in spirit.  Particles of phase 1 are not counted while evaluating density of phase 2 and vice versa. Updating of density is then based on the following discretized equations.
\begin{equation}
<\rho_{\alpha}^a>=\frac{\sum m_b w_{\alpha b} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} \left(h\right) +\sum \frac{m_j}{\rho_j} w_{\alpha j} \left(h\right)} \label{eq:gov-sph-d1}
\end{equation}
\begin{equation}
<\rho_\alpha^{sg}>=\frac{\sum m_j w_{\alpha j} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} \left(h\right) +\sum \frac{m_j}{\rho_j} w_{\alpha j} \left(h\right)} \label{eq:gov-sph-d2}
\end{equation}
where the subscript $a$ and $b$ represents air particles (phase 1) while $i$ and $j$ represents erupted material. $\beta$ and $\alpha$ represents either erupted material or air.
$\rho_a^a$ is density of phase 1 (air). 
 $\rho_i^{sg}$ is density of phase 2 (erupted material).
$\rho=\rho^a + \rho^{sg}$ is density of mixture of air and erupted material. 
By definition, the mass fraction is updated according to Eq. (\ref{eq:gov-sph-xi}).
\begin{equation}
<\xi_{\alpha}> = \dfrac{\rho^{sg}_{\alpha}}{\rho_{\alpha}}
\label{eq:gov-sph-xi}
\end{equation}

In areas far away from the interface, updating of density is exactly the same as that for single phase flow. For example, in the right side and left side (or blue areas) in Fig. \ref{fig:SPH-multiple-density}, where there are only air particles, Eq.  (\ref{eq:gov-sph-d2}) evaluates to zero and total density is:
\begin{equation}
\rho_{\alpha}=\rho_{\alpha}^a=\frac{\sum m_b w_{\alpha b} \left(h\right)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} \left(h\right)}
\end{equation}
Which is a special case of Eq. (\ref{eq:CSP-function-approximation-1d}). For these areas occupied by only particles of phase 2 and far away from the interface, similarly, the equation for density update becomes: 
\begin{equation}
\rho_{\alpha}=\rho_{\alpha}^{sg}=\frac{\sum m_j w_{\alpha j} \left(h\right)}{\sum \frac{m_j}{\rho_j} w_{\alpha j} \left(h\right)}
\end{equation}
That is to say, the same density updating equation can be applied for both phases and no additional numerical treatment needed to locate where the interface is.

\begin{figure}
\includegraphics[width=12cm]{Interface}
\caption{In the left figure, the blue particles (phaseID=1) represent air particles the red ones (phaseID=2) represent erupted material. The right figure shows corresponding mass fraction. Mass fraction are evaluated based on Eq. (\ref{eq:gov-sph-d1}) and Eq. (\ref{eq:gov-sph-d2}) without any other interface track or capture method.}
\label{fig:SPH-multiple-density}
\end{figure}

@@

More details about the data structure, domain decomposition, load balancing, domain adjusting and performance benchmarking has been published separately \citep{cao2017data}.
